A Transactional Memory System for Parallel Python

### 2014-06-07 talk/dls2014/paper/paper.tex
updates

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, they can not make use of the increasing parallelism in
current hardware.

In this paper, we present a software transactional memory (STM) system
specifically designed for the purpose of replacing the GIL in dynamic
language interpreters. The key idea is a close integration of STM with
the garbage collector to provide atomicity, synchronisation, and
memory management in a unified way.  As a user abstraction for a
scalable synchronisation mechanism, we implement atomic blocks by
exposing part of the STM system to applications.  The described STM
uses a combination of features present in current CPUs to lower its
overhead so that it can beat the performance of a GIL supported Python
interpreter already on 2 threads.

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is
therefore beneficial also for environments with a small number of
threads or cores. It is also used to implement atomic blocks for the
Python language to provide a backwards compatible, composable
synchronisation mechanism for threads.

@@

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism. At this point we should mention that it is indeed
possible to run external functions written in C instead of Python in
parallel. Our work focuses on Python itself and ignores this aspect as
it requires writing in a different language.

The second approach, \emph{multiprocessing}, uses multiple instances
of the interpreter itself and runs them in separate OS processes.
Here we actually get parallelism because we have one GIL per
interpreter, but of course we have the overhead of multiple processes~/
interpreters and also need to exchange data between them explicitly
and expensively.

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We follow the direct approach of using TM instead of the
GIL.

@@

Our TM system is fully implemented in software. However, we exploit
some more advanced features of current CPUs, particularly \emph{memory
segmentation, virtual memory,} and the 64-bit address space. Still,
it cannot be classified as a hybrid TM system since it currently
makes no use of any HTM present in the CPU.

@@

Clearly, if we are in a shared-memory setting, this model is extremely
wasteful: at any given time, most objects are identical in all, or
most, segments.  They differ only in segments that have modified but
not yet committed these objects.  So we use a trick of the Memory
Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
physical memory.  By changing this page mapping, at the coarse unit of
one page of memory (4096 bytes), we can share memory that would
otherwise exist as identical copies in all segments.

@@

The result is that the virtual address space of the process is bloated
by a factor $N$ -- but the real memory usage is not, because most
pages are shared with their identical copies in other segments.

@@

In summary, translating an $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments.

@@

  Checking other transactions' read marks would involve reading
  concurrently modified memory: another thread could be issuing a read
  barrier in parallel with a commit.  To prevent this from occurring, we
  must suspend all other threads at known safe points.  This is a
  trade-off for the simplicity of our read barrier -- which is well
  worth it in our case, due to the very high number of read barriers
  when compared to the number of commits.

@@

Conflicts can be resolved by aborting one of the two transactions, or
in some cases by pausing the writing one until the reader commits.  If a
commit succeeds, we copy the changes made to the objects in the write
set so that they become visible in other segments.  Note that this
only includes objects that existed before the transaction and have
been modified by it.  It does not include all the new objects created
by the current transaction, which typically outnumber modified objects
by far in dynamic languages.  We allocate these new objects in pages
that are still shared, so that they do not need to be copied to other
segments at all.

@@

In this section, we will provide details about the actual
implementation of the system and discuss some of the issues that we
encountered.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g.\ it shares the
write barrier).

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *SO)
void stm_write(object_t *SO)
object_t *stm_allocate(size_t size_rounded)
STM_PUSH_ROOT(object_t *SO)
STM_POP_ROOT(object_t *SO)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction, aborting it otherwise. \lstinline!stm_read()!,
\lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack locally around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

In the following sections, whenever we use $SO$, we go through the
address translation to get to the actual contents of an object. This
is also signified by the type \lstinline!object_t!.  This type is
special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system should use this type so
that the CPU will translate the reference to the right version of the
object.

\medskip   % why, Latex, why??
\subsection{Setup\label{sub:Setup}}

>>>>

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, applications can not make use of the increasing parallelism in
current hardware.

In this paper, we present a software transactional memory (STM) system
specifically designed for the purpose of replacing the GIL in dynamic
language interpreters. The key idea is a close integration of STM with
the garbage collector to approach atomicity, synchronisation, and
memory management in a unified way.  As a user abstraction for a
scalable synchronisation mechanism, we implement atomic blocks by
exposing part of the STM system to applications.  The described STM
uses a combination of features present in current CPUs, so that the
parallelisation benefits outweigh the system's overhead already on 2
threads when comparing to a GIL supported Python interpreter.

@@

This paper makes the following contributions:
\begin{itemize}[noitemsep]
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}
The STM described here is efficient and has low overhead and is
therefore beneficial also for environments with a small number of
threads or cores.

@@

\emph{Threading} employs operating system (OS) threads to provide
concurrency. It is, however, limited by the GIL and thus does not
provide parallelism\footnote{At this point we should mention that it is indeed
possible to run external functions written in C instead of Python in
parallel. Our work focuses on Python itself and ignores this aspect as
it requires writing in a different language.}. Because the GIL is only
seen as an implementation detail of the interpreter, applications are
required to coordinate concurrent accesses to data structures using
conventional methods -- locks being the most common way.

The second approach, \emph{multiprocessing}, uses multiple instances
of the interpreter itself and runs them in separate OS processes.
Here we actually get parallelism because there is one GIL per
interpreter, but of course we have the overhead of multiple processes~/
interpreters and also need to exchange data between them explicitly
and expensively.

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on multiple
threads. One approach to this is fine-grained locking instead of a
single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We on the other side follow the direct approach of using TM
instead of the GIL. There, GIL acquisition and release operations
conceptually directly map to starting and committing transactions.

TM also allows us to attack the issue of application-level
synchronisation and atomicity. We introduce atomic blocks to the
Python language by mapping them directly to the underlying
transactions -- providing the application with the guarantee of
atomicity for the enclosed instructions.

@@

The described TM system is fully implemented in software. However, we exploit
some more advanced features of current CPUs, particularly \emph{memory
segmentation, virtual memory,} and the 64-bit address space. Still,
it cannot be classified as a hybrid TM system since it currently
makes no use of any hardware TM (HTM) present in the CPU.

@@

Clearly, if we are in a shared-memory setting, this model is extremely
wasteful: at any given time, most objects are identical in all, or
most, segments.  They differ only in segments that have modified but
not yet committed these objects.  So we make use of the Memory
Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
physical memory.  By changing this page mapping, at the coarse unit of
one page of memory (4096 bytes), we can share memory that would
otherwise exist as identical copies in all segments.

@@

The result is that the virtual address space of the process is bloated
by a factor $N$ -- but the real, physical memory usage is not, because most
pages are shared with their identical copies in other segments.

@@

In summary, translating an $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments, automatically resolving either to a
shared or a private version.

@@

  Checking other transactions' read marks would involve reading
  concurrently modified memory: another thread could be issuing a read
  barrier in parallel to a commit.  To prevent this from occurring, we
  must suspend all other threads at known safe points.  This is a
  trade-off for the simplicity of our read barrier -- which is well
  worth it in our case, due to the very high number of read barriers
  when compared to the number of commits.

@@

Conflicts can be resolved by aborting one of the two transactions, or
in some cases by pausing the writing one until the reader commits --
serialising the write after the read.  If a
commit succeeds, we copy the changes made to the objects in the write
set so that they become visible in other segments.  Note that this
only includes objects that existed before the transaction and have
been modified by it.  It does not include all the new objects created
by the current transaction, which typically outnumber modified objects
by far in dynamic languages.  We allocate these new objects in pages
that are still shared, so that they do not need to be copied to other
segments at all.

@@

In this section, we will provide more details about the actual
implementation of the system and discuss some of the issues that we
encountered.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part.

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_become_inevitable()
void stm_read(object_t *SO)
void stm_write(object_t *SO)
object_t *stm_allocate(size_t size_rounded)
STM_PUSH_ROOT(object_t *SO)
STM_POP_ROOT(object_t *SO)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction, aborting it otherwise. Similarly,
\lstinline!stm_become_inevitable()!  tries to make the current
transaction inevitable, possibly aborting the current one if not
possible.
 \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack locally around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified Application
Programming Interface (API), only \lstinline!stm_allocate()!,
\lstinline!stm_become_inevitable()! and
\lstinline!stm_commit_transaction()!  require saving object
references.

In the following sections, whenever we use $SO$, we go through the
address translation to get to the actual contents of an object. This
is also signified by the type \lstinline!object_t!.  This type is
special as it causes the
compiler\footnote{We use Clang 3.5 with patches to deal with bugs in
its ``address-space 256'' feature. Patches are available from authors
until inclusion into the official clang.} to make all accesses through
it relative to the $\%gs$ register.  With exceptions, nearly all
accesses to objects managed by the TM system use this type so
that the CPU will translate the reference to the right version of the
object.
\medskip   % why, Latex, why??

\subsection{Replacing the GIL}

With this simple API, replacing the GIL is just a matter of starting a
new transaction when we would usually acquire the GIL, and committing
the current transaction when we normally release the GIL.  It would be
wasteful to do this between every single bytecode instruction.  We
thus depend on a simple heuristic that estimates how expensive
aborting and retrying the current transaction is based on how much
memory was allocated in the current transaction. If we deem it
expensive enough, we commit and start a new transaction \emph{as soon
as possible}. Deciding on the length of a transaction is an
interesting challenge in and of itself.

It may not be possible to immediately break the transaction, as we
could be executing an atomic block. Atomic blocks are currently
implemented as guaranteeing that the whole block's execution takes
place in a single transaction. Another issue is the support of
irreversible operations, like calls to external functions or general
input~/ output. If we are allowed to, we simply commit the transaction
before and start a new one after these operations. Within an atomic
block, however, we turn the transaction inevitable so that the
transaction cannot abort and strong isolation is guaranteed.

\subsection{Setup\label{sub:Setup}}
